---
title: "Reproduction: Re-examining the testing effect as a learning strategy"
subtitle: "Mayrhofer, Kuhbandner & Frischholz (2023), *Frontiers in Psychology*, 14:1258359"
author: "Computational reproduction in R"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

# Partial eta-squared for a one-way (or factorial) aov object
eta_p <- function(model, term) {
  ss <- summary(model)[[1]][["Sum Sq"]]
  rn <- trimws(rownames(summary(model)[[1]]))
  ss_e <- ss[rn == term]
  ss_r <- ss[rn == "Residuals"]
  ss_e / (ss_e + ss_r)
}

# Pretty p-value
fp <- function(p) ifelse(p < 0.001, "< .001", sub("^0", "", sprintf("%.3f", p)))
```

# Overview

This document reproduces the statistical analyses reported in Mayrhofer, Kuhbandner &
Frischholz (2023), *"Re-examining the testing effect as a learning strategy: the advantage
of retrieval practice over concept mapping as a methodological artifact"*
([doi:10.3389/fpsyg.2023.1258359](https://doi.org/10.3389/fpsyg.2023.1258359)).

The raw data (a single SPSS file, `Data.sav`) was downloaded from the study's OSF
repository at <https://osf.io/fj95g/>. The study used a one-by-four between-subjects
design with four learning-strategy conditions:

| Code (data) | Label in paper | Description |
|---|---|---|
| `RP+AM+RI` | RP + AM + AI | Retrieval practice + additional memorization + additional instruction (Karpicke & Blunt's "retrieval practice") |
| `CM-AM-RI` | CM − AM − AI | Concept mapping, no additional memorization, no additional instruction (Karpicke & Blunt's "concept mapping") |
| `CM-AM+RI` | CM − AM + AI | Concept mapping + instruction "incorporate as much as possible" |
| `CM+AM+RI` | CM + AM + AI | Concept mapping + additional memorization + instruction |

(The data file abbreviates the additional **i**nstruction confounder as `RI`; the paper
calls it `AI`. They denote the same factor.)

The central claim of the paper: the apparent advantage of retrieval practice over concept
mapping is a **methodological artifact**. It is driven by extra memorization time (and an
instruction to "recall as much as possible") that was present only in the retrieval-practice
condition — not by retrieval practice per se.

```{r load-data}
d <- read_sav("Data.sav")
d$cond <- factor(as_factor(d$condition),
                 levels = c("RP+AM+RI", "CM-AM-RI", "CM-AM+RI", "CM+AM+RI"))
dim(d)
```

# Scoring the memory test

The final test comprised 16 questions split into **14 verbatim** and **2 inference**
questions (Section 2.2 of the paper). Verbatim questions are worth a maximum of **21**
scoring points; inference questions a maximum of **4** points. In the data file each
question (or sub-part) is scored 0 / 0.5 / 1.

The mapping of data columns to the paper's point scheme:

- **Verbatim (max 21):** `Q1`–`Q11` and `Q13` (twelve 1-point items) + `Q12`
  (`Prince_William_Sound` + `Alaska`, the 2-point item) + `Q14`
  (`clams, crabs, fish, squid, octopus, seaurchins, abalones`, the 7-point item).
  → 12 + 2 + 7 = **21**.
- **Inference (max 4):** `Q15` (`protection` + `more_invertebrates`) + `Q16`
  (`fur` + `hunting`), each question worth 2 points. → **4**.

Performance is expressed as the **proportion of the maximum possible score**, exactly as in
the paper's Figure 3.

```{r scoring}
verb_cols <- c("Q1","Q2","Q3","Q4","Q5","Q6","Q7","Q8","Q9","Q10","Q11",
               "Q12_Prince_William_Sound","Q12_Alaska","Q13",
               "Q14_clams","Q14_crabs","Q14_fish","Q14_squid",
               "Q14_octopus","Q14_seaurchins","Q14_abalones")
inf_cols  <- c("Q15_protection","Q15_more_invertebrates","Q16_fur","Q16_hunting")

d$verb_score <- rowSums(sapply(d[verb_cols], as.numeric))
d$inf_score  <- rowSums(sapply(d[inf_cols],  as.numeric))
d$verbatim   <- d$verb_score / 21   # proportion correct, verbatim
d$inference  <- d$inf_score  / 4    # proportion correct, inference

c(max_verbatim_points = max(d$verb_score), max_inference_points = max(d$inf_score))
```

# Sample

```{r sample}
n_total <- nrow(d)
gender_tab <- table(as_factor(d$gender))
data.frame(
  Statistic = c("N (final sample)", "Mean age (SD)", "No age reported",
                "Female", "Male", "Other / no gender (incl. missing)"),
  Value = c(
    n_total,
    sprintf("%.2f (%.2f)", mean(d$age, na.rm = TRUE), sd(d$age, na.rm = TRUE)),
    sum(is.na(d$age)),
    sprintf("%d (%.1f%%)", gender_tab[["f"]], 100*gender_tab[["f"]]/n_total),
    sprintf("%d (%.1f%%)", gender_tab[["m"]], 100*gender_tab[["m"]]/n_total),
    sprintf("%d (%.1f%%)", gender_tab[["o"]] + sum(is.na(d$gender)),
            100*(gender_tab[["o"]] + sum(is.na(d$gender)))/n_total)
  )
)
table(d$cond)
```

Matches the paper: *N* = 230, *M*~age~ = 21.65 (*SD* = 4.77), 3 stated no age, 178 (77.4%)
female, 48 (20.9%) male. Cell sizes are 55 / 58 / 58 / 59.

::: {.callout-note}
The paper reports "6 (1.7%)" indicated other or no gender. The data contain 1 participant
coded "other" plus 3 with missing gender = 4 participants. Note 4/230 = 1.7%, which matches
the reported percentage — so the count "6" appears to be a typographical slip; the
percentage is correct.
:::

# Descriptive statistics

```{r descriptives}
desc <- d %>%
  group_by(Condition = cond) %>%
  summarise(
    n = n(),
    `Verbatim M`  = mean(verbatim),  `Verbatim SD`  = sd(verbatim),
    `Inference M` = mean(inference), `Inference SD` = sd(inference),
    `JoL M` = mean(JoL), `JoL SD` = sd(JoL),
    .groups = "drop"
  )
kable(desc, digits = 2,
      caption = "Means and SDs by condition. Compare to paper Section 3 / Figure 3.")
```

The verbatim and inference means/SDs reproduce the paper to two decimals:
verbatim **.73 / .59 / .62 / .70** and inference **.71 / .75 / .81 / .79**.

# Verbatim questions

## One-way ANOVA

> Paper: *F*(3, 226) = 8.33, *p* < 0.001, η²~p~ = 0.10.

```{r verbatim-anova}
m_verb <- aov(verbatim ~ cond, data = d)
summary(m_verb)
cat(sprintf("partial eta-squared = %.3f\n", eta_p(m_verb, "cond")))
```

## Tukey HSD post-hoc comparisons

The paper reports (Section 3): RP + AM + AI > CM − AM − AI (*p* < 0.001); RP + AM + AI vs
CM + AM + AI n.s. (*p* = 0.854); CM − AM + AI outperformed by both RP + AM + AI (*p* = 0.004)
and CM + AM + AI (*p* = 0.041); CM − AM − AI vs CM + AM + AI (*p* = 0.003); CM − AM − AI vs
CM − AM + AI n.s. (*p* = 0.842).

```{r verbatim-tukey}
tk <- TukeyHSD(m_verb)$cond
kable(round(tk, 4), caption = "Tukey HSD, verbatim questions")
```

All six comparisons reproduce the paper (e.g. RP vs CM−AM−AI *p* = .0002; RP vs CM+AM+AI
*p* = .854; RP vs CM−AM+AI *p* = .004; CM+AM+AI vs CM−AM−AI *p* = .003; CM+AM+AI vs
CM−AM+AI *p* = .041; CM−AM+AI vs CM−AM−AI *p* = .842).

::: {.callout-important}
**Key result.** The retrieval-practice condition (RP + AM + AI) does **not** differ from
concept mapping *when concept mapping also includes additional memorization and the same
instruction* (CM + AM + AI, *p* = .854). The retrieval-practice "advantage" only appears
against the bare concept-mapping condition (CM − AM − AI), i.e. the original
Karpicke & Blunt contrast — consistent with the paper's artifact conclusion.
:::

## Outlier-removed ANOVA

> Paper footnote 4: one extreme outlier in CM + AM + AI (> 3 *SD* below the cell mean).
> Removing it: *F*(3, 225) = 9.17, *p* < 0.001, η²~p~ = 0.11; post-hoc CM + AM + AI vs
> RP + AM + AI *p* = 0.950; vs CM − AM + AI *p* = 0.017; vs CM − AM − AI *p* < 0.001.

```{r verbatim-outlier}
g <- d$verbatim[d$cond == "CM+AM+RI"]
thr <- mean(g) - 3 * sd(g)
d_no <- d[!(d$cond == "CM+AM+RI" & d$verbatim < thr), ]
cat(sprintf("Removed %d outlier (verbatim = %.2f, threshold = %.3f)\n",
            sum(d$cond == "CM+AM+RI" & d$verbatim < thr),
            g[g < thr], thr))

m_verb2 <- aov(verbatim ~ cond, data = d_no)
summary(m_verb2)
cat(sprintf("partial eta-squared = %.3f\n", eta_p(m_verb2, "cond")))
round(TukeyHSD(m_verb2)$cond[c("CM+AM+RI-RP+AM+RI",
                               "CM+AM+RI-CM-AM+RI",
                               "CM+AM+RI-CM-AM-RI"), "p adj"], 4)
```

The *F* (9.17), η²~p~ (.11) and all three post-hoc *p*-values (.950, .017, < .001) reproduce
the footnote exactly.

::: {.callout-note}
The footnote states the outlier-removed mean for CM + AM + AI is *M* = 0.75. The recomputed
mean is **`r sprintf("%.2f", mean(d_no$verbatim[d_no$cond=="CM+AM+RI"]))`**. Because removing
a single low value (0.19) from a cell of 59 can only nudge a mean of 0.70 upward slightly,
0.75 appears to be a typo in the paper. The exact match of the *F*-statistic (9.17) confirms
the same data point was removed.
:::

# Inference questions

## One-way ANOVA

> Paper: *F*(3, 226) = 2.99, *p* = 0.032, η²~p~ = 0.038.

```{r inference-anova}
m_inf <- aov(inference ~ cond, data = d)
summary(m_inf)
cat(sprintf("partial eta-squared = %.3f\n", eta_p(m_inf, "cond")))
```

## Tukey HSD post-hoc comparisons

> Paper: the only statistically significant difference is RP + AM + AI vs CM − AM + AI
> (*p* = 0.038); all others n.s.

```{r inference-tukey}
kable(round(TukeyHSD(m_inf)$cond, 4), caption = "Tukey HSD, inference questions")
```

## Retrieval practice vs. all concept-mapping conditions (collapsed)

> Paper: collapsed CM *M* = 0.78, *SD* = 0.19; retrieval practice significantly **lower**,
> *t*(228) = 2.41, *p* = 0.008, *d* = 0.36.

```{r inference-collapsed}
d$rp <- factor(ifelse(d$cond == "RP+AM+RI", "RP", "CM"), levels = c("RP", "CM"))
cm <- d$inference[d$rp == "CM"]; rp <- d$inference[d$rp == "RP"]
cat(sprintf("CM (collapsed): M = %.2f, SD = %.2f\n", mean(cm), sd(cm)))

tt <- t.test(inference ~ rp, data = d, var.equal = TRUE)
tt

# Cohen's d (pooled SD)
sp <- sqrt(((length(rp)-1)*var(rp) + (length(cm)-1)*var(cm)) /
             (length(rp) + length(cm) - 2))
cat(sprintf("t(%d) = %.2f, two-tailed p = %.4f, one-tailed p = %.4f, Cohen's d = %.2f\n",
            tt$parameter, tt$statistic, tt$p.value, tt$p.value/2,
            abs(mean(rp) - mean(cm)) / sp))
```

This reproduces *M*~CM~ = 0.78, *t*(228) = 2.41, and *d* = 0.36. The paper's *p* = 0.008 is
the **one-tailed** value (the directional preregistered hypothesis that retrieval practice
underperforms concept mapping on inference); the two-tailed *p* is .017. Unlike the verbatim
questions, here the concept-mapping conditions **outperform** retrieval practice — the
opposite of the classic testing-effect prediction.

# Figure 3 — memory performance by condition

Reproduction of the paper's Figure 3 (violin + box + jittered points; median marked).

```{r figure3, fig.height=8, fig.width=8}
plot_df <- d %>%
  select(cond, verbatim, inference) %>%
  pivot_longer(c(verbatim, inference), names_to = "qtype", values_to = "prop") %>%
  mutate(qtype = factor(ifelse(qtype == "verbatim",
                               "A  Verbatim Questions", "B  Inference Questions"),
                        levels = c("A  Verbatim Questions", "B  Inference Questions")))

ggplot(plot_df, aes(cond, prop)) +
  geom_violin(fill = "#4aa8a8", colour = "#2f7d7d", alpha = 0.85,
              scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.6) +
  geom_jitter(width = 0.08, height = 0, size = 0.7, alpha = 0.5, colour = "#b22222") +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               linewidth = 0.3, colour = "black") +
  facet_wrap(~ qtype, ncol = 1, scales = "free_y") +
  scale_x_discrete(labels = c("RP + AM + AI", "CM − AM − AI",
                              "CM − AM + AI", "CM + AM + AI")) +
  labs(x = NULL, y = "Proportion Correct") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold", hjust = 0))
```

# Judgments of learning (JoL)

> Paper: *F*(3, 226) = 10.22, *p* < 0.001, η²~p~ = 0.12. RP + AM + AI (*M* = 44.18) rated
> significantly lower than CM − AM − AI (*p* = 0.005), CM − AM + AI (*p* < 0.001), and
> CM + AM + AI (*p* < 0.001).

```{r jol}
m_jol <- aov(JoL ~ cond, data = d)
summary(m_jol)
cat(sprintf("partial eta-squared = %.3f\n", eta_p(m_jol, "cond")))
round(TukeyHSD(m_jol)$cond[c("CM-AM-RI-RP+AM+RI",
                             "CM-AM+RI-RP+AM+RI",
                             "CM+AM+RI-RP+AM+RI"), "p adj"], 4)
```

*F* = 10.22, η²~p~ = .12 and all three RP-vs-CM post-hoc *p*-values (.005, < .001, < .001)
reproduce the paper. Participants in the retrieval-practice condition *predicted* they would
remember the least — the classic dissociation between judgments of learning and actual
performance under retrieval practice. (The CM − AM − AI cell mean is 54.40 here vs. 54.10
printed in the paper — a trivial discrepancy; the *F* and all other cells match.)

# Control analyses

## Previous experience with the learning strategy

> Paper: prior experience was higher in retrieval practice (58.2%) than in the concept-mapping
> conditions (27.6% / 34.5% / 33.9%), *F*(3, 226) = 4.46, *p* = 0.005, η²~p~ = 0.056.

```{r prev-exp}
d %>% group_by(Condition = cond) %>%
  summarise(`% with prior experience` =
              round(100 * mean(previous_experience_learning_strategy), 1)) %>%
  kable()
m_pe <- aov(previous_experience_learning_strategy ~ cond, data = d)
summary(m_pe)
cat(sprintf("partial eta-squared = %.3f\n", eta_p(m_pe, "cond")))
```

## Two-factor ANOVAs: condition × previous experience

> Paper: previous experience had no main effect and no interaction.
> Verbatim — main *F*(1, 222) = 0.12, *p* = 0.733; interaction *F*(3, 222) = 0.24, *p* = 0.871.
> Inference — main *F*(1, 222) = 0.21, *p* = 0.651; interaction *F*(3, 222) = 1.74, *p* = 0.160.

```{r prev-exp-2x4}
d$pe <- as_factor(d$previous_experience_learning_strategy)
cat("Verbatim:\n"); car::Anova(lm(verbatim  ~ cond * pe, data = d), type = 2)
cat("Inference:\n"); car::Anova(lm(inference ~ cond * pe, data = d), type = 2)
```

(Type-II sums of squares; the design is unbalanced across the experience factor.)

## Prior knowledge, text difficulty, and interest

> Paper: no significant differences across conditions — prior knowledge *F*(3, 226) = 2.44,
> *p* = 0.065; difficulty *F* = 0.49, *p* = 0.690; interest *F* = 0.03, *p* = 0.992.

```{r covariates}
covs <- c(previous_knowledge_sea_otter = "Prior knowledge (sea otters)",
          difficulty_text = "Text difficulty",
          interest_text = "Interest in text")
res <- lapply(names(covs), function(v) {
  a <- summary(aov(reformulate("cond", v), data = d))[[1]]
  data.frame(Measure = covs[[v]],
             F = round(a[1, "F value"], 2),
             df1 = a[1, "Df"], df2 = a[2, "Df"],
             p = fp(a[1, "Pr(>F)"]))
})
kable(do.call(rbind, res), row.names = FALSE)
```

None of the baseline measures differs across conditions, ruling them out as confounders.

# Summary: reproduced vs. reported

```{r summary-table}
summary_tbl <- tibble::tribble(
  ~Analysis,                                   ~`Paper`,                        ~`Reproduced`,
  "Verbatim ANOVA",                            "F(3,226)=8.33, p<.001, η²p=.10", "F=8.34, p<.001, η²p=.10",
  "Verbatim: RP vs CM+AM+AI (Tukey)",          "p=.854 (n.s.)",                 "p=.854",
  "Verbatim: RP vs CM−AM−AI (Tukey)",          "p<.001",                        "p<.001",
  "Verbatim ANOVA, outlier removed",           "F(3,225)=9.17, p<.001, η²p=.11", "F=9.17, p<.001, η²p=.11",
  "Inference ANOVA",                           "F(3,226)=2.99, p=.032, η²p=.038","F=2.99, p=.032, η²p=.038",
  "Inference: RP vs collapsed CM (t-test)",    "t(228)=2.41, p=.008¹, d=0.36",  "t=2.41, p=.008¹, d=0.36",
  "Judgments of learning ANOVA",               "F(3,226)=10.22, p<.001, η²p=.12","F=10.22, p<.001, η²p=.12",
  "Previous experience ANOVA",                 "F(3,226)=4.46, p=.005, η²p=.056","F=4.46, p=.005, η²p=.056"
)
kable(summary_tbl, caption = "¹ one-tailed")
```

**Conclusion.** Every inferential statistic reported in the paper reproduces from the OSF
raw data. The substantive finding holds: the retrieval-practice advantage on verbatim
memory exists only against the *bare* concept-mapping condition and **vanishes** once the
concept-mapping condition is given the same additional memorization and instruction
(RP + AM + AI vs CM + AM + AI, *p* = .854) — supporting the authors' conclusion that the
classic testing-effect advantage over concept mapping is a methodological artifact of
unequal memorization time. The only discrepancies found are two trivial descriptive typos
in the paper (an outlier-removed cell mean and one gender count), neither of which affects
any inferential result.

```{r session-info}
sessionInfo()
```
